Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2020-10-26 paper/main.tex
write conclusions

In many practical optimization problems, the objective function is not
explicitly available and solutions are evaluated by means of expensive
prediction models, simulations or physical experiments. When decision variables
are continuous, the use of Bayesian surrogate-models (e.g., Gaussian processes)
is widespread in
optimization~\citep{JonSchWel98go,ForKea2009surrogate}. Motivated by this
success, there have been attempts at adapting such Bayesian optimization
algorithms to the combinatorial case, a notable example being Combinatorial
Efficient Global Optimization
(CEGO)~\citep{ZaeStoBar2014:ppsn,ZaeStoFriFisNauBar2014}. However, the
ruggedness of combinatorial landscapes, which makes local search particularly
effective in the white-box context, lessens the effectiveness of global
surrogate models~\citep{EriPeaGar2019scalable}. Moreover, surrogate models are
expensive to train and optimize. The additional time required by the Bayesian
optimizer may impose a significant overhead in computation time, specially if
computation time of each function evaluation is measured in ``few'' minutes or
hours rather than days and when ``expensive'' refers to resources or economical
cost rather than wall-clock time. \citet{PerLopStu2015si} recently showed that
ant colony optimization (ACO)~\citep{DorStu2004:book} is competitive with CEGO
on a black-box version of the travelling salesman problem under a budget of
$1\,000$ function evaluations.  On the other hand, $1\,000$ evaluations is
still a relatively large budget for CEGO, whereas ACO was not designed for such
short budgets.

@@

Let us assume a black-box ``fitness'' function that must be minimized over a
space of permutations $f\colon S_n \to \mathbb{R}$. Being black-box means that
we can evaluate any candidate permutation $\sigma \in S_n$ to obtain
$f(\sigma)$, however, we do not know anything else about the form of
$f$. Moreover, the evaluation of $f$ is expensive in computation time or
resources and, thus, we can only evaluate a limited budget of candidate
permutations. Here, we will study budgets lower than 400. Such expensive
black-box combinatorial problems arises in diverse contexts, for example,
protein folding~\citep{RomKraArn2012protein} and industrial
production~\citep{FerAlvDiaIglEna2014ants}, where the fitness function may
involve expensive simulations for which no closed-form mathematical description
is available.

@@

\subsubsection{Learning step.}
At each iteration we start by fitting a MM to the sample $S$, which implies
estimating $\hat\sigma_0$ and $\theta$. Parameter $\theta$ is updated
deterministically at the sampling step.  The parameter $\sigma_0$, however, is
learned from the current sample $S$ by using the uBorda algorithm with a weight
$w(\sigma)=\rho^{f(\sigma)}$ (Section~\ref{sec:uborda}). The learning rate is
controlled by parameter $\rho \in (0,1)$. %
In UMM, the value of $\rho$ depends on the set of true fitness evaluations $F$
and it is set at each iteration such that the largest $r_1$\% of the mass of
the weights is concentrated in the best $r_2$\% of the solutions in $S$, i.e.,
\MANUEL{Let's talk about this Equation tomorrow}
%
\begin{equation}
\begin{split}
r_1 \sum_{\sigma\in S}\rho^{f(\sigma)} =  \sum_{\sigma\in S'}\rho^{f(\sigma)} \quad\text{where}\quad |S'| = r_2|S| \land \forall \sigma'\in S', \sigma\in S\setminus S' : f(\sigma')\geq f(\sigma)
\end{split}
\end{equation}
%
\MANUEL{should the $\sigma$ be the same on both sides of $=$?}
where $S'$ contains the $r_2|S|$ elements of $S$ with the best corresponding fitness values in $F$. The values of $r_1, r_2 \in (0, 1)$ are user-defined and control the balance between exploration and intensification: higher values bias the sampling towards the best solutions found so far, which may lead to quick stagnation, whereas smaller values make the sampling more uniform, thus slowing down the learning.

@@

\subsubsection{Other}.  For simplicity, we use Kendall's $\tau$
distance~(Eq.~\ref{eq:kendall_decomp}) in all
experiments\MANUEL{\citet{ZaeStoBar2014:ppsn} claims this is called Swap
  distance, could you double-check?}. Nevertheless,
\citet{ZaeStoFriFisNauBar2014} points out that CEGO performs best on the QAP
when using Hamming distance.  We plan to extend UMM to other distance measures
and to the dynamic selection of distance measures~\citep{ZaeStoBar2014:ppsn}.
In all experiments, we consider a maximum budget of $\FEmax=400$ evaluations of
the actual objective function. In a white-box context, state-of-the-art
algorithms for the problems considered here typically evaluate
\textcolor{red}{hundreds? thousands?} of solutions, thus, the budget considered
here for the black-box context is extremely limited.

@@

@@

The first part of the experimental analysis concerns the tunning of parameter $\rho$. As stated in Section~\ref{sec:sampling}, $\rho$ is automatically set at each iteration in such a way that the 100$r_1\%$ of the samples have the 100$r_2\%$ of the weight. In this regard, we tried different configurations of parameters $r_1$ and $r_2$ in all the instances. Figure~\ref{fig:heatmaps} shows the best solution (averaged for 10 runs) for instances: \texttt{N-t1d100.01} and \texttt{N-t1d150.01} for the PFSP, \texttt{reC05} for the PFSP, \texttt{kra32} for the QAP.

\begin{figure}[H]
  \centering%
  \includegraphics[width=0.24\textwidth]{../img/heatmap_N-t1d100_01}
  \includegraphics[width=0.24\textwidth]{../img/heatmap_N-t1d150_01}
  \includegraphics[width=0.24\textwidth]{../img/heatmap_rec05}
  \includegraphics[width=0.24\textwidth]{../img/heatmap_kra32_dat}
  \label{fig:heatmaps}
  \caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on LOP synthetic instances.\label{fig:lop_synth}}
\end{figure}

Overall the instances we see that the configuration $(r_1, r_2)=(0.1, 0.9)$ has good performance. Therefore, for the lack consistency along the experimental part, we use this configuration. Moreover, we recommend this configuration as a first approach with UMM.

Moreover, we can see that the behavior of the  two left figures differs from the two in the right: while the two in the left appear to be increase homogeneously from the top right corner to the bottom left one, the two in the right appear to be closer to randomness. We point out that this performance is related to the adequacy of UMM to the problem: homogeneousness is associated to the problem (not to the particular instances) and the more homogeneous the plot, the better performance of UMM. We elaborate this argument in the following sections with detail.

@@

%: LOP reales: he puesto 3 de diferentes tamanos. Me ha parecido interesante que, a pesar de que CEGO parece que va mejor, esta diferencia se va reduciendo segun se aumenta el tamano del problema (que se puede ver en el nombre del problema y en el títutlo de la gráfica)
\begin{figure}[tb]
  \centering%
  \includegraphics[width=\textwidth]{../img/fitness_real_lop_RandA1_N-t1d100_01}\\
  \includegraphics[width=\textwidth]{../img/fitness_real_lop_RandA1_N-t1d150_01}\\
  \includegraphics[width=\textwidth]{../img/fitness_real_lop_RandA1_N-t1d200_01}\\
    \caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on three instances from LOPLIB.\label{fig:loplib}}
  \end{figure}

@@

\begin{figure}
  \centering%
  \includegraphics[width=\textwidth]{../img/fitness_real_qap_kra32_dat}
    \caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on the QAP instance \texttt{kra32}.\label{fig:kra32}}
\end{figure}

@@

Bayesian optimization methods using a global GP model, such as CEGO, are known
to have trouble optimizing locally \citep{EriPeaGar2019scalable}. Our
intuition is that this problem becomes worse in rugged combinatorial
landscapes, where small steps may produce drastic changes.

Open questions:
\begin{itemize}
\item How difficult is to extend UMM to other distance metrics?
\item Can we plot the posterior probability of the optimal solution?
\end{itemize}

\begin{smaller}
\paragraph*{Acknowledgements.}
M.\@ L\'opez-Ib\'a\~nez is a ``Beatriz Galindo'' Senior Distinguished Researcher (BEAGAL 18/00053) funded by the Ministry of Science and Innovation of the Spanish Government.
%
Thanks to Hao Wang (Leiden University) for pointing us to the arguments of
\citet{EriPeaGar2019scalable}.
\end{smaller}

>>>>

In many practical optimization problems, the objective function is not
explicitly available and solutions are evaluated by means of expensive
prediction models, simulations or physical experiments. When decision variables
are continuous, the use of Bayesian surrogate-models (e.g., Gaussian processes)
is widespread in
optimization~\citep{JonSchWel98go,ForKea2009surrogate}. Motivated by this
success, there have been attempts at adapting such Bayesian optimization
algorithms to the combinatorial case, a notable example being Combinatorial
Efficient Global Optimization
(CEGO)~\citep{ZaeStoBar2014:ppsn,ZaeStoFriFisNauBar2014}. However, the
ruggedness of combinatorial landscapes, which makes local search particularly
effective in the white-box context, lessens the effectiveness of global
surrogate models~\citep{EriPeaGar2019scalable}. Moreover, surrogate models are
expensive to train and optimize. The additional time required by the Bayesian
optimizer may impose a significant overhead in computation time, specially if
computation time of each function evaluation is measured in ``few'' minutes or
hours rather than days and when ``expensive'' refers to resources or economical
cost rather than wall-clock time. \citet{PerLopStu2015si} recently showed that
ant colony optimization (ACO) %~\citep{DorStu2004:book}
is competitive with CEGO
on a black-box version of the travelling salesman problem under a budget of
$1\,000$ function evaluations.  On the other hand, $1\,000$ evaluations is
still a relatively large budget for CEGO, whereas ACO was not designed for such
short budgets.

@@

Let us assume a black-box ``fitness'' function that must be minimized over a
space of permutations $f\colon S_n \to \mathbb{R}$. Being black-box means that
we can evaluate any candidate permutation $\sigma \in S_n$ to obtain
$f(\sigma)$, however, we do not know anything else about the form of
$f$. Moreover, the evaluation of $f$ is expensive in computation time or
resources and, thus, we can only evaluate a limited budget of candidate
permutations. Here, we will study budgets lower than 400. Such expensive
black-box combinatorial problems arise in diverse contexts, for example,
protein folding~\citep{RomKraArn2012protein} and industrial
production~\citep{FerAlvDiaIglEna2014ants}, where the fitness function may
involve expensive simulations for which no closed-form mathematical description
is available. The black-box formulation and limited budget preclude most of the
optimization techniques that are successful in combinatorial optimization, such
as constructive and perturbation heuristics and local search, which makes this
a particularly challenging scenario, perhaps even more than its continuous
counter-part.

@@

\subsubsection{Learning step.}
At each iteration we start by fitting a MM to the sample $S$, which implies
estimating $\hat\sigma_0$ and $\theta$. Parameter $\theta$ is updated
deterministically at the sampling step.  The parameter $\sigma_0$, however, is
learned from the current sample $S$ by using the uBorda algorithm with a weight
$w(\sigma)=\rho^{f(\sigma)}$ (Section~\ref{sec:uborda}). The learning rate is
controlled by parameter $\rho \in (0,1)$. %
In UMM, the value of $\rho$ depends on the set of true fitness evaluations $F$
and it is set at each iteration such that the largest $100r_1$\% of the mass of
the weights is concentrated in the best $100r_2$\% of the solutions in $S$, i.e.,
\MANUEL{Let's talk about this Equation tomorrow}
%
\begin{equation}
\begin{split}
r_1 \sum_{\sigma\in S}\rho^{f(\sigma)} =  \sum_{\sigma\in S'}\rho^{f(\sigma)} \quad\text{where}\quad |S'| = r_2|S| \land \forall \sigma'\in S', \sigma\in S\setminus S' : f(\sigma')\geq f(\sigma)
\end{split}
\end{equation}
%
\MANUEL{should the $\sigma$ be the same on both sides of $=$?}
where $S'$ contains the $r_2|S|$ elements of $S$ with the best corresponding fitness values in $F$. The values of $r_1, r_2 \in (0, 1)$ are user-defined and control the balance between exploration and intensification: higher values bias the sampling towards the best solutions found so far, which may lead to quick stagnation, whereas smaller values make the sampling more uniform, thus slowing down the learning.

@@

\subsubsection{Other}.  For simplicity, we use Kendall's $\tau$
distance~(Eq.~\ref{eq:kendall_decomp}) in all
experiments\MANUEL{\citet{ZaeStoBar2014:ppsn} claims this is called Swap
  distance, could you double-check?}. Nevertheless,
\citet{ZaeStoFriFisNauBar2014} points out that CEGO performs best on the QAP
when using Hamming distance.  We plan to extend UMM to other distance measures
and to the dynamic selection of distance measures~\citep{ZaeStoBar2014:ppsn}.
In all experiments, we consider a maximum budget of $\FEmax=400$ evaluations of
the actual objective function. In a white-box context, state-of-the-art
algorithms for the problems considered here typically evaluate
thousands of solutions, thus, the budget considered
here for the black-box context is extremely limited.

@@

\subsection{Analysis of $r_1$ and $r_2$ parameters}

@@

The first part of the experimental analysis concerns the \manuel{tunning}{analyses} of parameter $\rho$. As stated in Section~\ref{sec:sampling}, $\rho$ is automatically set at each iteration in such a way that the $100r_1\%$ of the samples have the $100r_2\%$ of the weight. In this regard, we tried different configurations of parameters $r_1$ and $r_2$ in all the instances.\MANUEL{why did we try this values.} Figure~\ref{fig:heatmaps} shows the best solution\MANUEL{best solution found at the end of a run?} (averaged for 10 runs) for instances: \texttt{N-t1d100.01} and \texttt{N-t1d150.01} for the LOP, \texttt{reC05} for the PFSP, \texttt{kra32} for the QAP. \MANUEL{The caption says synthetic instances}

\begin{figure}[htb]
  \centering%
  \includegraphics[width=0.24\textwidth]{../img/heatmap_N-t1d100_01}
  \includegraphics[width=0.24\textwidth]{../img/heatmap_N-t1d150_01}
  \includegraphics[width=0.24\textwidth]{../img/heatmap_rec05}
  \includegraphics[width=0.24\textwidth]{../img/heatmap_kra32_dat}
    \caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on LOP synthetic instances.\label{fig:heatmaps}}
\end{figure}

The plot shows that UMM is indeed very sensitive to the values of $r_1$ and
$r_2$, yet, low values of $r_1$ and high values of $r_2$ almost always produce
the best results, independently of the problem. Nevertheless, there are a few
cases where an interaction seems to occur and particular combinations perform
remarkably well.\MANUEL{why?}  In particular, we observe that the pattern on
the two left-most figures differs from the two right-most ones. While on the
left the fitness gets worse homogeneously from the top right corner to the
bottom left one, the pattern on the right is not as homogeneous, yet still
fitness gets worse with decreasing $r_2$.  We observe that this behavior is
related to the adequacy of UMM to the problem: the homogeneous behavior of
parameters $r_1$ and $r_2$ depends on the problem (not to the particular
instances) and the more homogeneous the behavior, the better performance of
UMM. We elaborate this argument in the following sections with detail.

Over all the instances \MANUEL{(which instances we tested? all synthetic?)} we
see that the configuration $r_1 =0.1$, $r_2=0.9$ has good
performance\MANUEL{how do you define this? The best mean?}. From this
preliminary experiment, it is not obvious how the best settings are related to
problem features, such as permutation length ($n$). On the other hand, given
the sharp transitions in the plots, it is clear that further fine-tuning may
improve performance and uncover further patterns.  Although automatic parameter
tuning may likely provide even more fine-tuned settings, our goal here is to
understand how UMM performs in comparison to an existing Bayesian method (CEGO)
and automatically tuning the parameters of CEGO is not computationally
feasible.  For consistency, in the following sections, we run UMM with this
parameter configuration. Moreover, we recommend this configuration as the
current default of UMM.

\subsection{Analysis of UMM and CEGO}

@@

%: LOP reales: he puesto 3 de diferentes tamanos. Me ha parecido interesante que, a pesar de que CEGO parece que va mejor, esta diferencia se va reduciendo segun se aumenta el tamano del problema (que se puede ver en el nombre del problema y en el títutlo de la gráfica)
\begin{figure}[tbp]
  \centering%
  \includegraphics[width=\textwidth]{../img/fitness_real_lop_RandA1_N-t1d100_01}\\
  \includegraphics[width=\textwidth]{../img/fitness_real_lop_RandA1_N-t1d150_01}\\
  \includegraphics[width=\textwidth]{../img/fitness_real_lop_RandA1_N-t1d200_01}\\
    \caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on three instances from LOPLIB.\label{fig:loplib}}
  \end{figure}

@@

\begin{figure}[tp]
  \centering%
  \includegraphics[width=\textwidth]{../img/fitness_real_qap_kra32_dat}
    \caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on the QAP instance \texttt{kra32}.\label{fig:kra32}}
\end{figure}

@@

In this paper, we have introduced UMM, a population-based probabilistic
algorithm based on an unbalanced Mallows model. The algorithm is designed for
black-box combinatorial problems on permutation landscapes and when the budget
of fitness evaluations is severely limited (up to $1000$ or less).  UMM is
specially well-suited for budget-limited scenarios that are still
time-sensitive, e.g., when the overhead incurred by the optimizer should be no
more than a few minutes added to each fitness evaluation. As shown by our
results, a Bayesian optimizer such as CEGO may sometimes require more than hour
per fitness evaluation. The time required to train the surrogate-model of CEGO
also sharply increases with permutation length, hence, UMM would be a
computationally feasible alternative for relatively large problem sizes.

Although we were inspired by previous work on ACO for expensive black-box
combinatorial problems~\citep{PerLopStu2015si}, UMM is, to the best of our
knowledge, the first EDA specifically designed for such problems that is able
to obtain competitive results when compared with Bayesian optimizers. The
results presented here show that, despite the intrinsic difficulty of this scenario, there are still significant advances yet to be made. Thus, we hope
that this paper will motivate further research. % on such problems.

A somewhat surprising result was that the behavior of UMM (and the same for
CEGO) on the PFSP is close to random. This result may be due to the distance
metric used, but further analysis is needed to corroborate this
conjecture. Here, we have focused on Kendall's $\tau$ distance, however, it is
possible to extend UMM to other distance metrics, which will allow us to
dynamically select among various distance metrics for an unknown black-box
permutation landscape~\citep{ZaeStoBar2014:ppsn}. Also, as shown by our
experiments, the behavior of the parameters ($r_1$ and $r_2$) of UMM is not
always obvious, thus a more detailed analysis would be needed to provide either
generally good static values or an online adaptation approach. Finally, we have
used here three very different combinatorial problems (LOP, PFSP and QAP) as
black-box benchmarks. However, an even more diverse range of problems would be
needed to understand the behavior of UMM on real-world black-box combinatorial
landscapes.

% Bayesian optimization methods using a global GP model, such as CEGO, are known
% to have trouble optimizing locally \citep{EriPeaGar2019scalable}. Our
% intuition is that this problem becomes worse in rugged combinatorial
% landscapes, where small steps may produce drastic changes.

% Open questions:
% \begin{itemize}
% \item How difficult is to extend UMM to other distance metrics?
% \item Can we plot the posterior probability of the optimal solution?
% \end{itemize}

%% MANUEL: Blind, so no acknowledgments yet.
\begin{smaller}
\paragraph*{Acknowledgements.}
% M.\@ L\'opez-Ib\'a\~nez is a ``Beatriz Galindo'' Senior Distinguished Researcher (BEAGAL 18/00053) funded by the Ministry of Science and Innovation of the Spanish Government.
%
Thanks to Hao Wang (Leiden University) for pointing us to the arguments of
\citet{EriPeaGar2019scalable}.
\end{smaller}
